---
title: "Bayesian models and Markov chains"
author: "Mohar Sen"
date: "6/14/2020"
output: pdf_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Research topic: Sleep deprivation

**Research Question**  
How does sleep deprivation impact reaction time?

**The Study**  

* measure reaction time on Day 0
* restrict sleep to 3 hours per night
* measure reaction time on Day 3
* measure the change in reaction time  

For subject $i$, let $Y_i$ be the change in reaction time (in ms) after 3 sleep deprived nights. Of course, people react differently to sleep deprivation. It's reasonable to assume that $Y_i$ are Normally distributed around some average $m$ with standard deviation $s$
  
$Y_i$ = change in reaction time(ms) for subject $i$

**Assume**  
$Y_i$ are Normally distributed around some average change in reaction time m with
standard deviation $s$.  

$Y_i \sim N(m, s^2)$

__Prior model for parameter $m$__  
$Y_i$ = change in reaction time (ms)  
$Y_i \sim N(m, s^2)$  
$m = average Y_i$ 

__Prior information:__

* with normal sleep, average reaction time is ~250 ms
* expect average to increase by ~50 m
* average is unlikely to decrease & unlikely to increase by more than ~150 ms

Thus, m ~ N(50, 25^2)

Also, 
* s>0
* with normal sleep, s.d. in reaction times is ~30 ms
* s is equally likely to be anywhere from 0 to 200 ms

Thus, s ~ Unif(0,200)

Therefore,
$Y_i \sim N(m, s^2)$
$m \sim N(50, 25^2)$
$s \sim Unif(0,200)$

```{r}
library(ggplot2)
```


## Normal-Normal priors
In the first step of your Bayesian analysis, you'll simulate the following prior models for parameters m and s: 

+ Use rnorm(n, mean, sd) to sample 10,000 draws from the m prior. Assign the output to prior_m.
+ Use runif(n, min, max) to sample 10,000 draws from the s prior. Assign the output to prior_s.
+ After storing these results in the samples data frame, construct a density plot of the prior_m samples and a density plot of the prior_s samples.


```{r}
# Take 10000 samples from the m prior
prior_m <- rnorm(n=10000, mean=50, sd=25)

# Take 10000 samples from the s prior    
prior_s <- runif(n=10000, min=0, max=200)

# Store samples in a data frame
samples <- data.frame(prior_m, prior_s)

# Density plots of the prior_m & prior_s samples    
ggplot(samples, aes(x = prior_m)) + 
    geom_density()
ggplot(samples, aes(x = prior_s)) + 
    geom_density()
```
The distributions of these random samples approximate the features of your Normal prior for m and Uniform prior for s.





