---
title: "Multivariate and generalized linear models"
author: "Mohar Sen"
date: "6/15/2020"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 4 goals
+ Incorporate categorical predictors into Bayesian models
+ Engineer multivariate Bayesian regression models
+ Extend our methodology for Normal regression models to generalized linear models: Poisson regression

For starters, let's load the RailTrail dataframe from mosaic package

```{r}
library(rjags)
library(ggplot2)
library(mosaicData)
data(RailTrail)
```

`RailTrail` contains data collected by the Pioneer Valley Planning Commission on the usage of a local rail-trail. For each of 90 days, they recorded the rail-trail `volume` (number of users) and whether it was a `weekday` (TRUE if yes and FALSE otherwise).

+ Confirm that `weekday` is recorded as a `factor` variable.
+ Construct density plots of weekday `volume` and weekend `volume` on the same frame.

```{r}
# Confirm that weekday is a factor variable
class(RailTrail$weekday)

# Construct a density plot of volume by weekday
ggplot(RailTrail, aes(x = volume, fill = weekday)) + 
    geom_density(alpha = 0.5)
```

## RJAGS simulation with categorical variables

Consider the Normal regression model of volume $Y_i$ by weekday status $X_i$:

+ likelihood: $Y_i \sim N(m_i, s^2)$ where $m_i=a+bX_i$
+ priors: $a \sim N(400,100^2)$, $b\sim  N(0,200^2)$, $s \sim Unif(0,200)$


You explored the relationship between $Y_i$ and $X_i$ for the 90 days recorded in RailTrail (in your workspace). In light of these data and the priors above, you will update your posterior model of this relationship. This differs from previous analyses in that $X_i$ is categorical. In rjags syntax, its coefficient b is defined by two elements, `b[1]` and `b[2]`, which correspond to the weekend and weekday levels, respectively. For reference, `b[1]` is set to 0. In contrast, `b[2]` is modeled by the prior for `b`.

**Instructions 1/3**

DEFINE your Bayesian model.

+ Define the likelihood model of `Y[i]` given `m[i]` and s where `m[i] <- a + b[X[i]]`. Note the new notation `b[X[i]]` here!
+ Specify the priors for `a`, `b` (via `b[1]` and `b[2]`), and `s`.
+ Store the model string as rail_model_1.

```{r}
# DEFINE the model    
rail_model_1 <- "model{
    # Likelihood model for Y[i]
    for(i in 1:length(Y)) {
      Y[i] ~ dnorm(m[i], s^(-2))
      m[i] <- a + b[X[i]]
    }

    # Prior models for a, b, s
    a ~ dnorm(400, 100^(-2))
    b[1] <- 0
    b[2] ~ dnorm(0, 200^(-2))
    s ~ dunif(0, 200)
}"
```

**Instructions 2/3**

COMPILE rail_model_1 using jags.model():

+ Establish a `textConnection()` to `rail_model_1`.
+ Provide the observed `RailTrail` data.
+ Specify a starting random number generating seed of 10.
+ Store the output in a jags object named `rail_jags_1`.

```{r}
# COMPILE the model
rail_jags_1 <- jags.model(
  textConnection(rail_model_1),
  data = list(Y = RailTrail$volume, X = factor(RailTrail$weekday)),
  inits = list(.RNG.name = "base::Wichmann-Hill", .RNG.seed = 10)
)
```

Note: the compilation step was throwing an error, which was solved by the response as shown in this stackoverflow question.

**Instructions 3/3**

+ SIMULATE a sample of 10,000 draws from the posterior model of `a`, `b`, and `s`. Store this mcmc.list as `rail_sim_1`.
+ Store the Markov chain output in a data frame named `rail_chains_1`.
`+ `plot()` the posterior samples in `rail_sim_1`.

```{r}
# SIMULATE the posterior    
rail_sim_1 <- coda.samples(model = rail_jags_1, variable.names = c("a", "b", "s"), n.iter = 10000)

# Store the chains in a data frame
rail_chains_1 <- data.frame(rail_sim_1[[1]])

# PLOT the posterior
plot(rail_sim_1)
```

**Interpreting coefficients**

```{r}
summary(rail_sim_1)
```


We see that typically, there are ~428 trail users on a weekend day and ~78 fewer users on a weekday. Exact number may vary depending on the time of generation of this document.

## Inference for volume by weekday

You will utilize these simulation data to make inferences about weekday trail volume.

**Instructions**

+ Combine the `a` and `b.2`. chain values to construct a chain of 10,000 values for the typical weekday trail volume. Store this as `weekday_mea`n in `rail_chains_1`.
+ Use `ggplot()` to construct a density plot of the `weekday_mean` chain values.
+ Construct a 95% credible interval for the typical weekday trail volume.

```{r}
# Construct a chain of values for the typical weekday volume
rail_chains_1 <- rail_chains_1 %>% 
    mutate(weekday_mean = a + b.2.)

# Construct a density plot of the weekday chain
ggplot(rail_chains_1, aes(x = weekday_mean)) + 
    geom_density()

# 95% credible interval for typical weekday volume
quantile(rail_chains_1$weekday_mean, probs=c(0.025, 0.975))
```

## Multivariate Bayesian regression

In your previous work, you observed that rail-trail `volume` tends to be lower on a `weekday` than a weekend. Some of the variability in `volume` might also be explained by outside temperature. For example, we might expect trail volume to increase on warm, pleasant days.

The `RailTrail` data set in your workspace includes `hightemp`, the observed high temperature (F) for each of the 90 days in the study period. You will use these data to explore the associations between trail `volume`, `weekday` status, and `hightemp`

**Instructions**

### Construct a scatterplot of volume by hightemp:

+ Use `color` to distinguish between weekdays & weekends.
+ Use `geom_smooth()` to highlight the linear relationship between the observed `volume` & `hightemp` values.

```{r}
# Construct a plot of volume by hightemp & weekday
ggplot(RailTrail, aes(y = volume, x = hightemp, color = weekday)) + 
    geom_point() + 
	geom_smooth(method = "lm", se = FALSE)
```


Notice that for the 90 days in the study period, volume tends to increase with temperature. Further, volume tends to be higher on weekends than on weekdays of the same temperature. 

### RJAGS simulation for multivariate regression

Consider the following Bayesian model of volume Y
i by weekday status Xi and temperature Z

Consider the Normal regression model of volume $Y_i$ by weekday status $X_i$:

+ likelihood: $Y_i \sim N(m_i, s^2)$ where $m_i=a+bX_i+cZ_i$
+ priors: $a \sim N(0,200^2)$, $b\sim  N(0,200^2)$, $c \sim N(0,20^2)$, $s \sim Unif(0,200)$

Your previous exploration of the relationship between `volume`, `weekday`, and `hightemp` in the `RailTrail` data provided some insight into this relationship. You will combine this with insight from the priors to develop a posterior model of this relationship using RJAGS. The `RailTrail` data are in your work space.

**Instructions 1/3**

DEFINE your Bayesian model and store it as `rail_model_2`. Specifically, utilizing the `dnorm()` and `dunif()` rjags functions: 

+ For each of the 90 subjects $i$, define $m_i$ and the model of $Y_i$ given $m_i$ and $s$ using RJAGS notation. To this end, remember that `b[X[i]]` is the rjags equivalent of $bX_i$.
+ Specify the priors for `a`, `b`, `c`, and `s`.

```{r}
# DEFINE the model    
rail_model_2 <- "model{
    # Likelihood model for Y[i]
    for(i in 1:length(Y)) {
      Y[i] ~ dnorm(m[i], s^(-2))
      m[i] <- a + b[X[i]] + c*Z[i]
    }

    # Prior models for a, b, s
    a ~ dnorm(0, 200^(-2))
    b[1] <- 0
    b[2] ~ dnorm(0, 200^(-2))
    c ~ dnorm(0, 20^(-2))
    s ~ dunif(0, 200)
}"
```

**Instructions 2/3**

COMPILE `rail_model_2` using the `jags.model()` function with the following details:

+ For reproducible results, specify a starting seed of 10 for the `base::Wichmann-Hill` random number generator `inits`.
+ Run 1 chain and store the output in a jags object named `rail_jags_2`.

```{r}
rail_jags_2 <- jags.model(textConnection(rail_model_2), 
    data = list(Y = RailTrail$volume, X = factor(RailTrail$weekday), Z = RailTrail$hightemp), 
    inits = list(.RNG.name = "base::Wichmann-Hill", .RNG.seed = 10))
```

**Instructions 3/3**

+ SIMULATE a sample of 10,000 draws from the posterior model of `a`, `b`, `c`, and `s`. Store this `mcmc.list` as `rail_sim_2`.
+ Store the Markov chain output in a data frame named `rail_chains_2`.
+ `plot()` the posterior samples in `rail_sim_2`.

```{r}
# SIMULATE the posterior    
rail_sim_2 <- coda.samples(model = rail_jags_2, variable.names = c("a", "b", "c", "s"), n.iter = 10000)

# Store the chains in a data frame
rail_chains_2 = data.frame(rail_sim_2[[1]])

# PLOT the posterior
plot(rail_sim_2)
```

### Interpreting multivariate regression parameters

```{r}
summary(rail_sim_2)
```

The posterior mean of c indicates that for both weekends and weekdays, typical rail volume increases by ~5.4 users for every 1 degree increase in temperature. . Typical volume is ~50 less on weekdays than on weekends of the same temperature.

###Posterior inference for multivariate regression

Construct a scatterplot of `volume` by `hightemp`.

+ Use `color` to distinguish between weekdays & weekends.
+ Superimpose a `red` line that represents the posterior mean trend of the linear relationship between `volume` and `hightemp` for weekends: `m = a + c Z`
+ Superimpose a `turquoise3` line that represents the posterior mean trend of the linear relationship between `volume` and `hightemp` for weekdays: `m = (a + b.2.) + c Z`

```{r}
# Plot the posterior mean regression models
ggplot(RailTrail, aes(x = hightemp, y = volume, color = weekday))+
  geom_point()+
  geom_abline(intercept = mean(rail_chains_2$a), slope = mean(rail_chains_2$c), color = "red")+
  geom_abline(intercept = mean(rail_chains_2$a+rail_chains_2$b.2.), slope = mean(rail_chains_2$c), color = "turquoise3")
```

Posterior analysis suggests that there's a positive association between volume and temperature. Further, the typical weekday volume is less than that on weekends of the same temperature. 



